---
title: "Objective_1_pipeline"
format: html
editor: visual
---

## Objective_1_pipeline 

This is following the analysis presented in: https://www.microbiologyresearch.org/content/journal/mgen/10.1099/mgen.0.001525#tab2

```{r setwd_libraries}
setwd("~/Objective1_BEAST/Objective_1_pipeline")

library(rhierbaps)
library(ggtree)
library(phytools)
library(ape)
library(tidyverse)
library(tidysq)
library(seqinr)
library(BactDating)
library(ape)
library(ggtree)
library(dplyr)  
library(stringr)
library(reshape2)
library(tidyr)
library(lubridate)
library(coda)
```


```{r clades}
########################
### rhierBAPS clades ###
########################

### Run Rhierbaps code to determine clades
set.seed(1653)

snp.matrix <- load_fasta("Data/ERL06-3581_clean.core.fa")

hb.results <- hierBAPS(snp.matrix, max.depth = 3, n.pops = 20, quiet = TRUE, n.extra.rounds = Inf, assignment.probs=TRUE)
head(hb.results$partition.df)
write.csv(hb.results$partition.df, "Data/rhierbaps2.csv")


hb.results$lml.list
hb.results$cluster.assignment.prob
write.csv(hb.results$cluster.assignment.prob, "Data/rhierbaps2_cluster_assign_prob.csv")

### Then add clade information to Genome_summaries_144.csv - note 3 and 4 collapsed to 3 as Clade 1b

iqtree <- read.tree("Data/ERL06-3581_clean.core.tree")

gg <- ggtree(iqtree, layout = "circular", options(ignore.negative.edge=TRUE))
gg <- gg %<+% hb.results$partition.df
gg <- gg + geom_tippoint(aes(color = factor(`level 1`)))
gg

ggsave("Figures/Clades.svg", plot = gg, width = 7, height = 5)
```


```{r chainging headers}
#########################################
### adding info to fasta file headers ###
#########################################

myfasta <- seqinr::read.fasta (file = "Data/ERL06-3581_clean.core.fa", seqtype = "DNA", as.string = TRUE, set.attributes = FALSE)

### Header csv needs to be in same order as fasta headers 

# Function to sort the FASTA file by headers
sort_fasta_by_headers <- function(input_fasta, output_fasta) {
  
  # Step 1: Read the FASTA file
  fasta_data <- read.fasta(input_fasta, seqtype = "DNA", as.string = TRUE)
  
  # Step 2: Extract headers and sequences
  headers <- names(fasta_data)
  sequences <- unlist(fasta_data)
  
  # Step 3: Sort headers
  sorted_headers <- sort(headers)
  
  # Step 4: Reorder sequences based on sorted headers
  sorted_sequences <- sequences[match(sorted_headers, headers)]
  
  # Step 5: Write the sorted sequences to a new FASTA file
  output_fasta_connection <- file(output_fasta, "w")
  for (i in 1:length(sorted_headers)) {
    write(paste(">", sorted_headers[i], sep=""), output_fasta_connection)
    write(sorted_sequences[i], output_fasta_connection)
  }
  close(output_fasta_connection)
  
  cat("FASTA file sorted by headers and saved to", output_fasta, "\n")
}

### list of isolates to use - all must be present in fasta file and sorted by name
### Clade information from Rhierbaps should be added to csv file

subsetlist <- read.csv ("Data/Genome_summaries_144.csv", header=T) 
subsetlist <- subsetlist[order(subsetlist$Isolate.ID),]

### prepare new headers
headers <- paste(subsetlist$Isolate.ID,subsetlist$Automated_SBI, subsetlist$Island, subsetlist$Sample.type,subsetlist$Date, sep="|")

### subset file
my_fasta_sub <- myfasta [names (myfasta) %in% subsetlist$Isolate.ID]

### Sort fasta file by headers
write.fasta (sequences = my_fasta_sub, names = names (my_fasta_sub), file.out = "Data/STEC2.fasta")
input_fasta <- "Data/STEC2.fasta"
output_fasta <- "Data/sorted_STEC2.fasta"
sort_fasta_by_headers(input_fasta, output_fasta)


### add dates to headers
read_fasta("Data/sorted_STEC2.fasta") %>%
  mutate(name = headers) %>%
  {
    .x <- .
    write_fasta(.x$sq, .x$name, "Data/STEC_144.fasta")
  }

```

```{r clade_subset}
#######################################
### Then subset by Rhierbaps clades ###
#######################################

### Clade 1
### subset file

subsetlist_clade1<-subsetlist[subsetlist$RB_level.1b=="1",]
my_fasta_sub <- myfasta [names (myfasta) %in% subsetlist_clade1$Isolate.ID]

### prepare new headers
headers <- paste(subsetlist_clade1$Isolate.ID,subsetlist_clade1$Automated_SBI, subsetlist_clade1$Island, subsetlist_clade1$Sample.type,subsetlist_clade1$Date, sep="|")

### Sort fasta file by headers
write.fasta (sequences = my_fasta_sub, names = names (my_fasta_sub), file.out = "Data/STEC1.fasta")
input_fasta <- "Data/STEC1.fasta"
output_fasta <- "Data/sorted_STEC1.fasta"
sort_fasta_by_headers(input_fasta, output_fasta)


### add dates to headers
read_fasta("Data/sorted_STEC1.fasta") %>%
  mutate(name = headers) %>%
  {
    .x <- .
    write_fasta(.x$sq, .x$name, "Data/STEC_Clade1.fasta")
  }

### Clade 2
### subset file
subsetlist_clade2<-subsetlist[subsetlist$RB_level.1b=="2",]
my_fasta_sub <- myfasta [names (myfasta) %in% subsetlist_clade2$Isolate.ID]

### prepare new headers
headers <- paste(subsetlist_clade2$Isolate.ID,subsetlist_clade2$Automated_SBI, subsetlist_clade2$Island, subsetlist_clade2$Sample.type,subsetlist_clade2$Date, sep="|")

### Sort fasta file by headers
write.fasta (sequences = my_fasta_sub, names = names (my_fasta_sub), file.out = "Data/STEC2.fasta")
input_fasta <- "Data/STEC2.fasta"
output_fasta <- "Data/sorted_STEC2.fasta"
sort_fasta_by_headers(input_fasta, output_fasta)


### add dates to headers
read_fasta("Data/sorted_STEC2.fasta") %>%
  mutate(name = headers) %>%
  {
    .x <- .
    write_fasta(.x$sq, .x$name, "Data/STEC_Clade2.fasta")
  }

### Clade 3
### subset file

subsetlist_clade1<-subsetlist[subsetlist$RB_level.1b=="3",]
my_fasta_sub <- myfasta [names (myfasta) %in% subsetlist_clade1$Isolate.ID]

### prepare new headers
headers <- paste(subsetlist_clade1$Isolate.ID,subsetlist_clade1$Automated_SBI, subsetlist_clade1$Island, subsetlist_clade1$Sample.type,subsetlist_clade1$Date, sep="|")

### Sort fasta file by headers
write.fasta (sequences = my_fasta_sub, names = names (my_fasta_sub), file.out = "Data/STEC3.fasta")
input_fasta <- "Data/STEC3.fasta"
output_fasta <- "Data/sorted_STEC3.fasta"
sort_fasta_by_headers(input_fasta, output_fasta)


### add dates to headers
read_fasta("Data/sorted_STEC3.fasta") %>%
  mutate(name = headers) %>%
  {
    .x <- .
    write_fasta(.x$sq, .x$name, "Data/STEC_Clade3.fasta")
  }

```


### Then run IQTREE to get the treefile and make it a nwk file

```{bash, eval=false}
#!/bin/bash -e
#SBATCH --account massey03727
#SBATCH --job-name=iqtree # job name (shows up in the queue)
#SBATCH --output=iqtree.log
#SBATCH --time=02:00:00 # Walltime (HH:MM:SS)
#SBATCH --cpus-per-task=16 # number of cpus/threads, in our case 4
#SBATCH --mem=20GB            # Memory in MB

module purge

module load IQ-TREE/2.2.2.2-gimpi-2022a

# iqtree -s STEC_144.fasta -m MF

iqtree -s STEC_Clade1.fasta -m MF

iqtree -s STEC_Clade2.fasta -m MF

iqtree -s STEC_Clade3.fasta -m MF

```



```{r midpoint_root_the_trees}
#### Then get ML nwk tree files for each fasta file set and import into TempEst
#### Parse dates - set format for dates as yyy-MM-dd

#tree <- read.tree("STEC_144.fasta.treefile")
#mid_tree <- midpoint.root(tree)
#write.tree(mid_tree, file="STEC_144.fasta.nwk")


tree1 <- read.tree("Data/STEC_Clade1.fasta.treefile")
mid_tree1 <- midpoint.root(tree1)
write.tree(mid_tree1, file="Data/STEC_Clade1.fasta.nwk")


tree2 <- read.tree("Data/STEC_Clade2.fasta.treefile")
mid_tree2 <- midpoint.root(tree2)
write.tree(mid_tree2, file="Data/STEC_Clade2.fasta.nwk")


tree3 <- read.tree("Data/STEC_Clade3.fasta.treefile")
mid_tree3 <- midpoint.root(tree3)
write.tree(mid_tree3, file="Data/STEC_Clade3.fasta.nwk")
```


```{r bactdating_clade1}
# As previously seen with the analysis from TempEST the core SNP phylogeny

############################## Clade 1 #####################################

#This tree has 50 isolates included. 
## read in tree
tree=read.tree("Data/STEC_Clade1.fasta.nwk")
#### make sure tip names and dates are in the same order ###
names<-data.frame(tree$tip.label)
header<-"name"
colnames(names) <-header

meta <- separate(data = names, 
                 col = name, 
                 into = c("name", "SBI", "island", "host", "date"), 
                 sep = "\\|") ## needs // as | special char 

dates=decimal_date(as.Date(meta$date))

#make sure branch lengths are number of substitutions and not per site 
## Check - should be total length of all branchesand much more than 1
sum(tree$edge.length)
#[1] 0.2588016 = not 1, need to do next step


## if not do this, where multiplyer is length of snp profile
### but make sure reload TREE before repeating this line!!! ###

# SNP profile length is 3698 - I looked in geneious
tree$edge.length=tree$edge.length*3698

tree=initRoot(tree, dates, mtry = 10, useRec = F)

## sense check
allDates(tree)
## parameter estimates
roottotip(tree,dates, showFig=F)

## complete plot
roottotip(tree,dates, showFig=T)

## just root to tip correlation
r2<-roottotip(tree,dates,showFig=T,colored=T,showPredInt='gamma', showText=F,showTree=F)


# Alternative model
## fit bactdate model on time stamped tree
res=bactdate(tree,dates, model="arc", nbIts = 90000000, showProgress = T)
## check trace and convergence
plot(res,'trace')
mcmc=as.mcmc.resBactDating(res)
effectiveSize(mcmc) ## should be >200 if coverged
#      mu    sigma    alpha 
# 201.1904 501.0000 226.8020 
# Converged

# Null model
## fit bactdate model on fixed date tree (null model)
Res2= bactdate(tree,rep(2010,length(dates)), model="arc", nbIts = 5000000, showProgress = T)
## check trace and convergence
plot(Res2,'trace')
mcmc=as.mcmc.resBactDating(Res2)
effectiveSize(mcmc)
#         mu      sigma      alpha 
# 3.987165 413.810878  10.863473 
# Not converged

modelcompare(res,Res2)
#The first model has DIC=2169.81 and the second model has DIC=3675.40.
# Model 1 is definitely better.

```
> roottotip(tree,dates, showFig=F)
$rate
     date 
0.6643144 

$ori
(Intercept) 
  -83.35468 

$pvalue
[1] 0.0041

> roottotip(tree,dates, showFig=T)
$rate
     date 
0.6643144 

$ori
(Intercept) 
  -83.35468 

$pvalue
[1] 0.0043


> # Alternative model
> ## fit bactdate model on time stamped tree
> res=bactdate(tree,dates, model="arc", nbIts = 90000000, showProgress = T)
  |==============================================================================| 100%
  
> mcmc=as.mcmc.resBactDating(res)
> effectiveSize(mcmc) ## should be >200 if coverged
      mu    sigma    alpha 
166.0129 501.0000 146.1781 

```{r bactdating_clade2}
############################## Clade 2 #####################################

# This tree has 46 isolates included. 
## read in tree
tree2=read.tree("Data/STEC_Clade2.fasta.nwk")
#### make sure tip names and dates are in the same order ###
names2<-data.frame(tree2$tip.label)
header2<-"name"
colnames(names2) <-header2

meta2 <- separate(data = names2, 
                  col = name, 
                  into = c("name", "SBI", "island", "host", "date"), 
                  sep = "\\|") ## needs // as | special char 

dates2=decimal_date(as.Date(meta2$date))

#make sure branch lengths are number of substitutions and not per site 
## Check - should be total length of all branchesand much more than 1
sum(tree2$edge.length)
# [1] 0.2455159 again not 1 


## if not do this, where multiplyer is length of snp profile
### but make sure reload TREE before repeating this line!!! ###

# SNP profile length is 3698 - I looked in geneious
tree2$edge.length=tree2$edge.length*3698

tree2=initRoot(tree2, dates2, mtry = 10, useRec = F)

## sense check
allDates(tree2)
## parameter estimates
roottotip(tree2,dates2, showFig=F)

## complete plot
roottotip(tree2,dates2, showFig=T)

## just root to tip correlation
r22<-roottotip(tree2,dates2,showFig=T,colored=T,showPredInt='gamma', showText=F,showTree=F)


# Alternative model
## fit bactdate model on time stamped tree
res22=bactdate(tree2,dates2, model="arc", nbIts = 5000000, showProgress = T)
## check trace and convergence
plot(res22,'trace')
mcmc=as.mcmc.resBactDating(res22)
effectiveSize(mcmc) ## should be >200 if coverged
#    mu sigma alpha 
# 501   501   501  = converged

# Null model
## fit bactdate model on fixed date tree (null model)
Res22= bactdate(tree2,rep(2010,length(dates2)), model="arc", nbIts = 5000000, showProgress = T)
## check trace and convergence
plot(Res22,'trace')
mcmc=as.mcmc.resBactDating(Res22)
effectiveSize(mcmc)
#        mu      sigma      alpha 
# 3.356791 588.733071  20.059068 

modelcompare(res22,Res22)
#The first model has DIC=487.59 and the second model has DIC=6238.22.
# Model 1 is definitely better.
```

> ## parameter estimates
> roottotip(tree2,dates2, showFig=F)
$rate
    date 
1.515432 

$ori
(Intercept) 
   -11.7541 

$pvalue
[1] 0.0327

> ## complete plot
> roottotip(tree2,dates2, showFig=T)
$rate
    date 
1.515432 

$ori
(Intercept) 
   -11.7541 

$pvalue
[1] 0.0329

> # Alternative model
> ## fit bactdate model on time stamped tree
> res22=bactdate(tree2,dates2, model="arc", nbIts = 5000000, showProgress = T)
  |=====================================================================================================================| 100%
> ## check trace and convergence
> plot(res22,'trace')
> mcmc=as.mcmc.resBactDating(res22)
> effectiveSize(mcmc) ## should be >200 if coverged
       mu     sigma     alpha 
 11.44074 501.00000  14.71377 
> ## check trace and convergence
> plot(res22,'trace')
> # Null model
> ## fit bactdate model on fixed date tree (null model)
> Res22= bactdate(tree2,rep(2010,length(dates2)), model="arc", nbIts = 5000000, showProgress = T)
  |=====================================================================================================================| 100%
> ## check trace and convergence
> plot(Res22,'trace')
> mcmc=as.mcmc.resBactDating(Res22)
> effectiveSize(mcmc)
        mu      sigma      alpha 
  9.485197 501.000000   7.413765 

```{r bactdating_clade3}
############################## Clade 3 ##################################### 

# This tree has 48 isolates included. 
## read in tree
tree3=read.tree("Data/STEC_Clade3.fasta.nwk")
#### make sure tip names and dates are in the same order ###
names3<-data.frame(tree3$tip.label)
header3<-"name"
colnames(names3) <-header3

meta3 <- separate(data = names3, 
                  col = name, 
                  into = c("name", "SBI", "island", "host", "date"), 
                  sep = "\\|") ## needs // as | special char 

dates3=decimal_date(as.Date(meta3$date))

#make sure branch lengths are number of substitutions and not per site 
## Check - should be total length of all branchesand much more than 1
sum(tree3$edge.length)
# [1] 0.4515492 nope not 1

## if not do this, where multiplyer is length of snp profile
### but make sure reload TREE before repeating this line!!! ###

# SNP profile length is 3698 - I looked in geneious
tree3$edge.length=tree3$edge.length*3698

tree3=initRoot(tree3, dates3, mtry = 10, useRec = F)

## sense check
allDates(tree3)
## parameter estimates
roottotip(tree3,dates3, showFig=F)
#$rate
#date 
#1.255433 

#$ori
#(Intercept) 
#1887.482 

#$pvalue
#[1] 0.0104

## complete plot
roottotip(tree3,dates3, showFig=T)
#$rate
#date 
#1.255433 

#$ori
#(Intercept) 
#1887.482 

#$pvalue
#[1] 0.0099
## just root to tip correlation
r23<-roottotip(tree3,dates3,showFig=T,colored=T,showPredInt='gamma', showText=F,showTree=F)

# Alternative model
## fit bactdate model on time stamped tree
res23=bactdate(tree3,dates3, model="arc", nbIts = 100000000, showProgress = T) ### increase the chain length ### up to 10 mil or so
## check trace and convergence
plot(res23,'trace')
mcmc=as.mcmc.resBactDating(res23)
effectiveSize(mcmc) ## mu should be >200 if coverged
# mu     sigma     alpha 
# 231.46059 702.82088  55.76344 
 


# Null model
## fit bactdate model on fixed date tree (null model)
Res23= bactdate(tree3,rep(2010,length(dates3)), model="arc", nbIts = 100000000, showProgress = T)
## check trace and convergence
plot(Res23,'trace')
mcmc=as.mcmc.resBactDating(Res23)
effectiveSize(mcmc)
#         mu      sigma      alpha 
# 3.270484 501.000000  16.716363 

modelcompare(res23,Res23)
# The first model has DIC=1243.65 and the second model has DIC=368638.94.
# Model 1 is definitely better.
```

